Resilience and Home-Space for WSTS \thanks{This work was partly done while the authors were supported by the Agence Nationale de la Recherche grant BraVAS (ANR-17-CE40-0028).}

### 2023-10-15 VMCAI-revised-version/final.tex
effectivity

We introduce five types of effectiveness \cite{BFM-ic17} that will be essential for solving various resilience problems in different models of WSTS.
\noindent
%			{\bf On effectivity.}
We say that a WSTS $\mathscr{S}=(S, \rightarrow, \leq)$ is 
\begin{itemize}
\item {\em effective} if there exists a pair of algorithms
($M_\rightarrow$, $M_\leq$) operating on $\N \times \N$ such that
$ M_\rightarrow$ computes the transition relation “$\rightarrow$” and 
$M_\leq$ the ordering relation “$\leq$”.
\item {\em post-effective} if $\mathscr{S}$ is effective, and if there
exists an algorithm that computes $|\post(x)| \in \N \cup \{\infty\}$ 
on input $x \in X$. 
\item {\em effective pred-basis} \cite{DBLP:journals/tcs/FinkelS01,DBLP:journals/iandc/AbdullaCJT00} if there exists an algorithm accepting
any state $s \in S$ and returning $pb(s)$, a finite basis of $\uparrow \pred(\uparrow s)$.
%
\item {\em ideally-effective} \cite{BFM-ic17} if (1) the function mapping the encoding of a state $s$
to the encoding of the ideal $\downarrow s$ is computable; (2) inclusion of ideals is decidable; (3) the downward closure $\downarrow \post(I)$ expressed as a finite union of ideals is computable from the ideal $I$.
%
\item {\em completion-post-effective} if $\mathscr{S}$ is ideally-effective and its completion $\hat{\mathscr{S}}=(Ideals(S),\rightarrow, \subseteq)$ is post-effective.
%
%and there exist an algorithm $M_\downarrow$ that computes 
% $\downarrow s$, on input $s \in S$, and some algorithm $M_{\uparrow^C}$ that computes the ideal decomposition of $S \setminus \uparrow \{ s_1, s_2, \ldots, s_m\}$, on input
% $s_1, s_2, \ldots, s_m \in S$.
%
\end{itemize}
\alain{le rappel ci-dessous est-il entièrement utile ? Je pense qu'on peut l'enlever, et toi ?}
Now, we may recall a simple condition that insures that a finite basis of $\pred^*(\uparrow s )$ is computable for every $s \in S$.
We will use the following property: if $\mathscr{S}$ is a WSTS with strong compatibility and $U \subseteq S$ is upward-closed, then $\pred(U )$, $\pred^k(U )$ with $k\geq0$, and $\pred^*(U )$ are all upward-closed \cite{DBLP:journals/tcs/FinkelS01}.
For a WSTS $\mathscr{S}=(S, \rightarrow, \leq)$ and an upward-closed set $U  \subseteq S$, let us study the convergence of the sequence defined by $U_0=U$ and $U_k= U_{k-1} \cup \pred(U_{k-1})$ for $k \geq 1$. When $\mathscr{S}$ has strong compatibility, the sets $U_k$ are upward-closed and $U_k \subseteq U_{k+1}$ so we know that the sequence $(U_k)_k$ converges. Let us define the \emph{index} of convergence of the sequence $U_k$ as the smallest $k_0$ s.t. $U_k = U_{k_0}$ for all $k \geq k_0$. We may compute $k_0$ and we then have:  $\pred^*(U) = U_{k_0}$. When $\mathscr{S}$ have upward-compatibility but not strong compatibility, we can similarly compute $\pred^*(U)$ by studying the sequence $(\uparrow U_k)_k$ instead.
With the effective pred-basis hypothesis, we obtain:

@@

Let us recall that a Well Behaved Transition System (WBTS) \cite{DBLP:journals/lmcs/BlondinFM17} is an upward compatible ordered transition system $\mathscr{S}=(S, \rightarrow, \leq)$ where $(S,\leq)$ that contains no infinite antichains (but it can be not well founded). The class of WBTS is strictly larger than WSTS: for example, integer VASS under the lexicographical ordering are WBTS but not WSTS \cite{DBLP:journals/lmcs/BlondinFM17}. 
With the ideal-effective hypothesis, we obtain the decidability of coverability for WBTS.

\begin{theorem}\cite{DBLP:journals/lmcs/BlondinFM17}
Coverability is decidable for any ideal-effective WBTS.
\end{theorem}

@@

\begin{theorem} \cite{BFM-ic17}
Coverability is  decidable for completion-post-effective WSTS.
\end{theorem}

@@

\begin{theorem}\label{indec WSTS}
Resilience for post-effective $\omega^2$-WSTS with strong compatibility and for upward closed set $\Safe$ is undecidable.
\end{theorem}
\alain{peut-on ajouter des hypothèses d'effectivity ? peut-on montrer que l'exemple n'est pas pred effectif ? l'exemple est-il  ideal-effective (oui pour les propriétes 1 et 2 mais 3 ?)? on voit que l'exemple n'est pas completion-post-effective vu le theoreme qui suit}
\begin{proof}
Indeed, consider the family $\{ f_j: \N^2 \to \N^2\}$ of 
increasing recursive functions from \cite{FMP-ic04} defined as

@@

	\noindent	
	where TM$_j$ is the $j$-th Turing machine (in a classical enumeration)
	which moreover begins
by writing the integer $j$ on its tape,	
	and consider additionnally the function $g : \N^2 \to \N^2$ defined by $g(n,k) = (n+1,k)$. 
	The
	transition system 
	$S_j = (\N^2, \{f_j, g\}, \leq)$ 
	is a post-effective $\omega^2$-WSTS with strong compatibility
	and
	has the property that
	it is
	$\uparrow (1,1)$-resilient
	iff TM$_j$ halts on input $j$,
	since $(1,1)$ is coverable from $(0,0)$
	iff TM$_j$ halts on input $j$. 
	Hence there is no Turing machine which correctly determines 
	resilience and halts whenever its input is a
	WSTS.
Hence resilience for $\omega^2$-WSTS with strong compatibility is undecidable.\qed
\end{proof}
 Let us remark that the transition system 
	$S_j = (\N^2, \{f_j, g\}, \leq)$ in the previous proof is not ideally effective neither completion-post-effective.

@@

%
\begin{theorem}\label{down-up}
Let $\mathscr{S}=(S,\rightarrow, \leq)$ be a completion-post-effective WSTS with strong compatibility and a set $\Safe = \uparrow \Safe$.
{\sc Resilience}, {\sc Bounded resilience} 
and {\sc $k$-resilience} are decidable.
\end{theorem}
\alain{a-t-on le même rsultat pour les WSTS avec pred* calculable ? J'ai enlevé l'hypothèse omega2 car j'ai l'impression qu'elle n'est pas utile...à vérifier}
\begin{proof}
Let us first recall two results in \cite{BFM-ic17}. Proposition 30 establishes a strong relation between the runs of a WSTS $\mathscr{S}=(S,\rightarrow, \leq)$ and the runs of its completion $\hat{\mathscr{S}}$. It states that if $x \xrightarrow{k} y$ in $\mathscr{S}$ then for every ideal $I \supseteq \downarrow x$, there exists an ideal $J \supseteq \downarrow y$ such that $I \xrightarrow{k} J$ in $\hat{\mathscr{S}}$. Proposition 29 establishes that if $I \xrightarrow{k} J$ in $\hat{\mathscr{S}}$ then for every $y \in J$, there exists $x \in I$ and $y' \geq y$ such that $x \xrightarrow{k'} y'$ in $\mathscr{S}$. Moreover, if $\mathscr{S}$ has transitive compatibility then $k’ \geq k$; if $\mathscr{S}$ has strong compatibility then $k’ = k$.
%

@@

This implies that resilience and bounded resilience are equivalent to coverability.
Since coverability is decidable for completion-post-effective WSTS \cite{BFM-ic17}, we deduce that both the 
  resilience problem and the bounded resilience problem are decidable.

@@

\begin{proof}
\alain{est-ce sur qu'on peut rajouter l'hypothese "ideal-effective" . preuve pas facile à lire, à réécrire et à repenser}
This stems from the fact that it is undecidable for Minsky machines with more than one counter,
which can be simulated by WSTS.
Indeed, {\sc Resilience} is undecidable for Minsky machine with at least $3$ counters.
This is due to the undecidability of $2$-counter Minsky machine termination~\cite{Min61,Min67}.
From a $2$-counter Minsky machine $M$, one can construct a $3$-counter Minsky machine $M'$ 
such that machine $M$ terminates for all inputs iff machine $M'$ can reach $(0,0,0)$ from any input with at least $1$ on its third counter. We build $M'$ to simulate $M$ until it reaches a control-state indicative of termination, then lower the first two counters until they reach $0$, then, and only then, finally lower the third counter until it reaches $0$.
Remark that in the construction, from any input with at least $1$ on its third counter, it is not possible to reach $(1,0,0)$, $(0,1,0)$ or $(1,1,0)$.
Based on this construction, the problem of deciding whether the downward closed set $\downarrow (1,1,0)$ is reachable in a $3$-counter Minsky machine
from any input with at least $1$ on its third counter is undecidable. 
Hence {\sc Resilience} is undecidable for Minsky machines.

@@

%      	it is well known that reset-VASS are WSTS~\cite{dufourd1998reset}. 
Since reset-VASS can simulate executions of a Minsky machine, {\sc Resilience} is undecidable for reset-VASS and hence for $\omega^2$-WSTS as well.\alain{déjà dit avant en mieux omega2 et strong comp et post-effective mais reset-VASS apportent effective pred-basis....}\qed
\end{proof}

@@

\begin{theorem}\label{downward srp}
{\sc State-resilience} is decidable for ideally-effective WSTS with downward and upward compatibilities,
$\Safe = \downarrow \Safe$.
\end{theorem}

@@

\begin{corollary}\label{downward brp}
{\sc $k$-state-resilience} and {\sc bounded-state-resilience} are decidable for ideally-effective WSTS with strong downward compatibility and upward compatibility when
$\Safe = \downarrow \Safe$.
\end{corollary}

@@

In this section we study VASS. Since they are completion-post-effective WSTS, they inherit the decidability results for WSTS in the case 
$\Safe = \uparrow \Safe$. Lacking downward compatibility or a more relaxed hypothesis that
for all downward-closed set $D$, the set $\pred^*(D)$ is downward-closed, 
VASS do not inherit the decidability results for WSTS in the case $\Safe = \downarrow \Safe$.
In this section, we work to re-establish decidability results for VASS when $\Safe$ is downward-closed. We also extend decidability to resilience for semilinear sets rather than simply upward and downward-closed ones.

@@

We have extended the state-resilience problem introduced in~\cite{DBLP:journals/corr/PrasadZ16,DBLP:journals/corr/abs-2108-00889,DBLP:conf/gg/Ozkan22}.
We complemented previous results by providing some undecidability proofs for resilience and state-resilience in general. We also exhibited classes of WSTS with decidable resilience, namely, 
completion-post-effective $\omega^2$-WSTS with strong compatibility (in the case 
 $\Safe = \uparrow \Safe$) and 
downward-compatible
ideal-effective WSTS
(in the case 
 $\Safe = \downarrow \Safe$).

>>>>

We introduce five types of effectiveness \cite{BFM-ic17} that will be essential for solving various resilience problems in different models of WSTS.
\noindent
%			{\bf On effectivity.}
We say that a WSTS $\mathscr{S}=(S, \rightarrow, \leq)$ is 
\begin{itemize}
\item {\em effective} if there exists a pair of algorithms
($M_\rightarrow$, $M_\leq$) operating on $\N \times \N$ such that
$ M_\rightarrow$ computes the transition relation “$\rightarrow$” and 
$M_\leq$ the ordering relation “$\leq$”.
%
\item {\em post-effective} if $\mathscr{S}$ is effective, and if there
exists an algorithm that computes $|\post(x)| \in \N \cup \{\infty\}$ 
on input $x \in X$.
%
\item {\em ideally-effective} \cite{BFM-ic17} if the function mapping the encoding of a state $s$
to the encoding of the ideal $\downarrow s$ is computable and inclusion of ideals is decidable.
%
\item {\em completion-post-effective} if its completion $\hat{\mathscr{S}}=(Ideals(S),\rightarrow, \subseteq)$ is post-effective, i.e., that $\downarrow \post(I)$ is computable, expressed as a finite union of ideals, for all $I \in Ideals(S)$.
%
\item {\em effective pred-basis} \cite{DBLP:journals/tcs/FinkelS01,DBLP:journals/iandc/AbdullaCJT00} if there exists an algorithm accepting
any state $s \in S$ and returning $pb(s)$, a finite basis of $\uparrow \pred(\uparrow s)$.
%
%and there exist an algorithm $M_\downarrow$ that computes 
% $\downarrow s$, on input $s \in S$, and some algorithm $M_{\uparrow^C}$ that computes the ideal decomposition of $S \setminus \uparrow \{ s_1, s_2, \ldots, s_m\}$, on input
% $s_1, s_2, \ldots, s_m \in S$.
%
\end{itemize}
%Most of usual WSTS enjoy most of all effectivities.

\alain{le rappel ci-dessous est-il entièrement utile ? Je pense qu'on peut l'enlever, et toi ?}
Now, we may recall a simple condition that insures that a finite basis of $\pred^*(\uparrow s )$ is computable for every $s \in S$.
We will use the following property: if $\mathscr{S}$ is a WSTS with strong compatibility and $U \subseteq S$ is upward-closed, then $\pred(U )$, $\pred^k(U )$ with $k\geq0$, and $\pred^*(U )$ are all upward-closed \cite{DBLP:journals/tcs/FinkelS01}.
For a WSTS $\mathscr{S}=(S, \rightarrow, \leq)$ and an upward-closed set $U  \subseteq S$, let us study the convergence of the sequence defined by $U_0=U$ and $U_k= U_{k-1} \cup \pred(U_{k-1})$ for $k \geq 1$. When $\mathscr{S}$ has strong compatibility, the sets $U_k$ are upward-closed and $U_k \subseteq U_{k+1}$ so we know that the sequence $(U_k)_k$ converges. Let us define the \emph{index} of convergence of the sequence $U_k$ as the smallest $k_0$ s.t. $U_k = U_{k_0}$ for all $k \geq k_0$. We may compute $k_0$ and we then have:  $\pred^*(U) = U_{k_0}$. When $\mathscr{S}$ have upward-compatibility but not strong compatibility, we can similarly compute $\pred^*(U)$ by studying the sequence $(\uparrow U_k)_k$ instead.
With the effective pred-basis hypothesis, we obtain:

@@

Let us recall that a Well Behaved Transition System (WBTS) \cite{DBLP:journals/lmcs/BlondinFM17} is an upward compatible ordered transition system $\mathscr{S}=(S, \rightarrow, \leq)$ where $(S,\leq)$ that contains no infinite antichains (but it can be not well founded). The class of WBTS is strictly larger than WSTS: for example, integer VASS under the lexicographical ordering are WBTS but not WSTS \cite{DBLP:journals/lmcs/BlondinFM17}. 
With the ideally-effective hypothesis, we obtain the decidability of coverability for WBTS.

\begin{theorem}\cite{DBLP:journals/lmcs/BlondinFM17}
Coverability is decidable for any ideally-effective WBTS.
\end{theorem}

@@

\begin{theorem} \cite{BFM-ic17}
Coverability is  decidable for completion-post-effective and ideally-effective WSTS.
\end{theorem}

@@

\begin{theorem}\label{indec WSTS}
Resilience for post-effective and ideally effective $\omega^2$-WSTS with strong compatibility, for upward closed sets $\Safe$ is undecidable.
\end{theorem}
\alain{peut-on ajouter des hypothèses d'effectivity ? peut-on montrer que l'exemple n'est pas pred effectif ? l'exemple est-il  ideal-effective (oui pour les propriétes 1 et 2 mais pas 3 car ça permettrait de décider l'arrêt ?)? on voit que l'exemple n'est pas completion-post-effective vu le theoreme qui suit}
\begin{proof}
Indeed, consider the family $\{ f_j: \N^2 \to \N^2\}$ of 
increasing recursive functions from \cite{FMP-ic04} defined as

@@

	\noindent	
	where TM$_j$ is the $j$-th Turing machine (in a classical enumeration)
	which moreover begins
by writing the integer $j$ on its tape,	
	and consider additionnally the function $g : \N^2 \to \N^2$ defined by $g(n,k) = (n+1,k)$. 
	The
	transition system 
	$S_j = (\N^2, \{f_j, g\}, \leq)$ 
	is a post-effective $\omega^2$-WSTS with strong compatibility
	and
	has the property that
	it is
	$\uparrow (1,1)$-resilient
	iff TM$_j$ halts on input $j$,
	since $(1,1)$ is coverable from $(0,0)$
	iff TM$_j$ halts on input $j$. 
	Hence there is no Turing machine which correctly determines 
	resilience and halts whenever its input is a
	WSTS.
Hence resilience for $\omega^2$-WSTS with strong compatibility is undecidable.\qed
\end{proof}
 Let us remark that the transition system 
	$S_j = (\N^2, \{f_j, g\}, \leq)$ in the previous proof is not completion-post-effective.

@@

%
\begin{theorem}\label{down-up}
Let $\mathscr{S}=(S,\rightarrow, \leq)$ be a completion-post-effective and ideally-effective WSTS with strong compatibility and a set $\Safe = \uparrow \Safe$.
{\sc Resilience}, {\sc Bounded resilience} 
and {\sc $k$-resilience} are decidable.
\end{theorem}
\alain{a-t-on le même rsultat pour les WSTS avec pred* calculable ? J'ai enlevé l'hypothèse omega2 car j'ai l'impression qu'elle n'est pas utile...à vérifier}
\begin{proof}
Let us first recall two results in \cite{BFM-ic17}. Proposition 30 establishes a strong relation between the runs of a WSTS $\mathscr{S}=(S,\rightarrow, \leq)$ and the runs of its completion $\hat{\mathscr{S}}$. It states that if $x \xrightarrow{k} y$ in $\mathscr{S}$ then for every ideal $I \supseteq \downarrow x$, there exists an ideal $J \supseteq \downarrow y$ such that $I \xrightarrow{k} J$ in $\hat{\mathscr{S}}$. Proposition 29 establishes that if $I \xrightarrow{k} J$ in $\hat{\mathscr{S}}$ then for every $y \in J$, there exists $x \in I$ and $y' \geq y$ such that $x \xrightarrow{k'} y'$ in $\mathscr{S}$. Moreover, if $\mathscr{S}$ has transitive compatibility then $k’ \geq k$; if $\mathscr{S}$ has strong compatibility then $k’ = k$.
%

@@

This implies that resilience and bounded resilience are equivalent to coverability.
Since coverability is decidable for completion-post-effective and ideally-effective WSTS \cite{BFM-ic17}, we deduce that both the 
  resilience problem and the bounded resilience problem are decidable.

@@

\begin{proof}
\alain{est-ce sur qu'on peut rajouter l'hypothese "ideal-effective" ? oui. et completion-post-effective ?...... preuve pas claire, à réécrire}
This stems from the fact that it is undecidable for Minsky machines with more than one counter,
which can be simulated by WSTS.
Indeed, {\sc Resilience} is undecidable for Minsky machine with at least $3$ counters.
This is due to the undecidability of $2$-counter Minsky machine termination~\cite{Min61,Min67}.
From a $2$-counter Minsky machine $M$, one can construct a $3$-counter Minsky machine $M'$ 
such that machine $M$ terminates for all inputs iff machine $M'$ can reach $(0,0,0)$ from any input with at least $1$ on its third counter. We build $M'$ to simulate $M$ until it reaches a control-state indicative of termination, then lower the first two counters until they reach $0$, then, and only then, finally lower the third counter until it reaches $0$.
Remark that in the construction, from any input with at least $1$ on its third counter, it is not possible to reach $(1,0,0)$, $(0,1,0)$ or $(1,1,0)$.
Based on this construction, the problem of deciding whether the downward closed set $\downarrow (1,1,0)$ is reachable in a $3$-counter Minsky machine
from any input with at least $1$ on its third counter is undecidable. 
Hence {\sc Resilience} is undecidable for Minsky machines.

@@

%      	it is well known that reset-VASS are WSTS~\cite{dufourd1998reset}. 
Since reset-VASS can simulate executions of a Minsky machine, {\sc Resilience} is undecidable for reset-VASS and hence for $\omega^2$-WSTS as well.\alain{déjà dit avant en mieux omega2 et strong comp et post-effective mais reset-VASS apportent effective pred-basis, completion et ideally effective....}\qed
\end{proof}

@@

\begin{theorem}\label{downward srp}
{\sc State-resilience} is decidable for ideally-effective and completion-post-effective WSTS with downward and upward compatibilities,
$\Safe = \downarrow \Safe$.
\end{theorem}

@@

\begin{corollary}\label{downward brp}
{\sc $k$-state-resilience} and {\sc bounded-state-resilience} are decidable for ideally-effective and completion-post-effective WSTS with strong downward compatibility and upward compatibility when
$\Safe = \downarrow \Safe$.
\end{corollary}

@@

In this section we study VASS. Since they are ideally-effective and completion-post-effective WSTS, they inherit the decidability results for WSTS in the case 
$\Safe = \uparrow \Safe$. Lacking downward compatibility or a more relaxed hypothesis that
for all downward-closed set $D$, the set $\pred^*(D)$ is downward-closed, 
VASS do not inherit the decidability results for WSTS in the case $\Safe = \downarrow \Safe$.
In this section, we work to re-establish decidability results for VASS when $\Safe$ is downward-closed. We also extend decidability to resilience for semilinear sets rather than simply upward and downward-closed ones.

@@

We have extended the state-resilience problem introduced in~\cite{DBLP:journals/corr/PrasadZ16,DBLP:journals/corr/abs-2108-00889,DBLP:conf/gg/Ozkan22}.
We complemented previous results by providing some undecidability proofs for resilience and state-resilience in general. We also exhibited classes of WSTS with decidable resilience, namely, 
completion-post-effective and ideally-effective WSTS with strong compatibility (in the case 
 $\Safe = \uparrow \Safe$) and 
downward-compatible
ideally-effective WSTS
(in the case 
 $\Safe = \downarrow \Safe$).
